Density-dependent selection and the limits of relative fitness

### 2018-02-28 manuscript.tex
l

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage[authoryear,round]{natbib}
\usepackage{lineno}
\usepackage{graphicx} 
\usepackage{setspace}
\doublespacing

@@

In this appendix we derive Eq.~\eqref{eq:master}. We start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
\begin{equation}
\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i,
\end{equation}
which vary in relative magnitude depending on the propagule densities $l_i$. The first component, $\Delta_u n_i$, accounts for territories where only one focal propagule is present ($x_i=1$ and $x_j=0$ for $j\neq i$; $u$ stands for ``uncontested''). The proportion of territories where this occurs is $l_i e^{-L}$, and so 
\begin{equation}
\Delta_u n_i=Ul_i e^{-L}=m_i e^{-L}.
\end{equation}

@@

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types. To illustrate, suppose that two identical (same $b$, $c$ and $d$) types are present, one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$. Since $L$ is large, uncontested territories contribute negligibly to each type's growth. The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This is a serious pathology: any single-type population can be arbitrarily partitioned into rare and common identical subtypes, and the predicted growth or decline of any small group of individuals would depend on whether it is regarded as part of a rare or a common type.

This pathology arises if we naively use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ to calculate the effective means. In this case we have $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq. \eqref{eq:meanfieldr} is approximately $1/(L+1)$ for the rare type, and, since $l_1\ll 1$ and $L\gg 1$ in Eq.~\eqref{eq:deltr}, we have $\Delta_r n_1 \approx 1/(L+1)$. Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}}\approx L$ in Eq. \eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$. Thus, the identical rare type is erroneously predicted to decline in frequency.

The effective distributions $\tilde{q}$ and $\hat{q}$ ensure that the above consistency requirement is satisfied.

$\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).

%Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.

By contrast, Eq. \eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''. By neglecting the latter covariances, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations. The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.

Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$. If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.

We first calculate $\langle x_j \rangle_{\tilde{p}}$. Let $X=\sum_j x_j$ denote the total number of propagules in a territory and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$. Then, $\tilde{p}$ is the probability distribution for ${\mathbf x_i}$ conditional on the territory having received $X\geq 2$ propagules, where $X_i=X-1$ of them are nonfocal, giving
\begin{equation}
\tilde{p}({\mathbf x_i})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x_i}|X_i=X-1)
\end{equation}
Here the total number of propagules $X$ follows a Poisson distribution with mean $L$, and we have $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$. Thus,
\begin{align}
\langle x_j \rangle_{\tilde{p}}&=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j\nonumber\\
&=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(X) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=X-1)x_j.
\label{eq:raremonster1}
\end{align}
The inner sum over ${\mathbf x_i}$ is the mean number of propagules of a given nonfocal type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total, which is equal to $\frac{l_j}{L-l_i}(X-1)$. Thus, 
\begin{align}
\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{X=2}^{\infty} P(X) (X-1)\nonumber\\
&=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
\label{eq:meanxjrare}
\end{align}
where the last line follows from $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)X-\sum_{X=1}^{\infty}P(X)$.

@@

\subsection*{Competition when abundant}

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage[authoryear,round]{natbib}
\usepackage{lineno}
\usepackage{graphicx} 
\usepackage{setspace}
\doublespacing
\sloppy

@@

In this appendix we derive Eq.~\eqref{eq:master}. Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.

We start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
\begin{equation}
\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i,\label{eq:delt_decomp}
\end{equation}
which vary in relative magnitude depending on the propagule densities $l_i$. The first component, $\Delta_u n_i$, accounts for territories where only one focal propagule is present ($x_i=1$ and $x_j=0$ for $j\neq i$; $u$ stands for ``uncontested''). The proportion of territories where this occurs is $l_i e^{-L}$, and so 
\begin{equation}
\Delta_u n_i=Ul_i e^{-L}=m_i e^{-L}.
\end{equation}

@@

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types. To illustrate, suppose that two identical (same $b$, $c$ and $d$) types are present, one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$. Since $L$ is large, uncontested territories make up a negligible fraction of the total, and the rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This behavior is pathological: any single-type population can be arbitrarily partitioned into identical rare and common subtypes, but a given small group of individuals would be predicted to either grow or shrink depending on whether it viewed as part of the rare or common type.

The above pathology arises if we naively use $\tilde{p}$ and $\hat{p}$ to calculate the effective means. The right hand side of Eq. \eqref{eq:meanfieldr} is then approximately $1/(1+L)$, and since $l_1\ll 1$ and $L\gg 1$ we have $\Delta_r n_1 \approx 1/(1+L)$ in Eq.~\eqref{eq:deltr}. Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = L$ in Eq. \eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$. Thus, the identical rare type is  pathologically predicted to decline in frequency.

The effective distributions $\tilde{q}$ and $\hat{q}$ are devised to avoid this pathology. Let ${\mathbf x}$ denote the vector of propagule abundances $(x_1,\ldots,x_G)$ in a given territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ similarly denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$. The corresponding total propagule numbers are denoted $X=\sum_j x_j$ and $X_i=X-x_i$. Then, in territories where one focal propagule and at least one non-focal propagule were received, the effective distribution is defined by 
\begin{equation}
\tilde{q}({\mathbf x_i})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x_i}|X_i=X-1),
\end{equation}
where the total number of propagules $X$ follows a Poisson distribution with mean $L$. Similarly, in territories where more than one focal propagule was received, the effective distribution is defined by 
\begin{equation}
\hat{q}({x})=\sum_{X=2}^{\infty}P(X|x_i\geq 2) p({\mathbf x}|x_i\geq 2, X).
\end{equation}
In other words, $\tilde{q}$ and $\hat{q}$ make the approximation that the distribution for the total number of propagules per territory is the same in all territories. In reality, conditioning on one or more focal propagules being present does affect the distribution of $X$ in the corresponding subset of territories. Thus, for $\tilde{p}$ and $\hat{p}$, which are simply the corresponding conditional distributions of the actual dispersal distribution $p$, the mean propagule density is $L$ but $\langle X \rangle_{\tilde{p}}=L+1$.

%By contrast, Eq. \eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''. By neglecting the latter covariances, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations. The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.

%Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$. If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.

\subsection*{Calculating the effective means}

We first calculate $\langle x_j \rangle_{\tilde{p}}$.

and we have $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$
 Thus,
\begin{align}
\langle x_j \rangle_{\tilde{p}}&=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j\nonumber\\
&=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(X) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=X-1)x_j.
\label{eq:raremonster1}
\end{align}
The inner sum over ${\mathbf x_i}$ is the mean number of propagules of a given nonfocal type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total, which is equal to $\frac{l_j}{L-l_i}(X-1)$. Thus, 
\begin{align}
\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{X=2}^{\infty} P(X) (X-1)\nonumber\\
&=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
\label{eq:meanxjrare}
\end{align}
where the last line follows from $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)X-\sum_{X=1}^{\infty}P(X)$.

\subsection*{Limits of the approximation}

@@
